---
title: "Final Project: Sexual Education, Pregnancy Risk, & Contraceptive Use"
author: "Amel Attalla & Belen Zemas"
date: "2025-06-26"
params:
  number: 6
  purpose: "Final project for PUBH 6199: Visualizing Data with R"
format: 
  html:
    toc: true
    toc-location: right
    code-fold: true
    theme: minty
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
output-dir: docs
---


```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
    warning = FALSE,
    message = FALSE,
    fig.path = "figs/",
    fig.width = 7.252,
    fig.height = 4,
    comment = "#>",
    fig.retina = 3
)
```

## Introduction

In this project we aim to examine the relationship between the types of sexual health education that are taught in the United States, contraceptive knowledge and risk of unplanned pregnancies, as well as how that knowledge is applied to real life. What constitutes appropriate and effective sexual education continues to be an increasingly debated topic, we aim to shed light on how different approaches-such as abstinence-only vs. contraceptive focused curricula may correlate with public health outcomes. For this project we have decided to create visualizations that focus on these three questions: 

1.⁠ ⁠How does sex education type influence perceptions of pregnancy risk?

2.⁠ ⁠How does sex education type influence the method of contraceptive used?

3.⁠ ⁠How do sexual education type, perceived pregnancy risk, and contraceptive method interrelate in shaping reproductive decision-making?


## Data

We sourced our raw data set from the ICPSR website. Some of the key variables we are focusing on after cleaning the raw data are: contraceptive use, perceived risk of unplanned pregnancy, and types of sexual health education received. Cleaning the data set mainly consisted of narrowing down the variables from 340 to 12 keeping only the variables that we felt were relevant to our project/ the visualizations we wanted to make. The steps taken to wrangle and clean the original data set can be seen in it's entirety in the code below.

```{r}
#| eval: false
# Load libraries and data
library(tidyverse)
library(dplyr)
library(haven)

raw_data <- read_dta("data/Guttmacher 2009 Survey of contraceptive knowledge (Fog Zone).dta")

clean_fogzone <- raw_data %>% 
  select(gender,age4grps,race,riskr,abstonly,protect,curmethrr,pregknowr,totknowr,nummthdsr,wprotect,mprotect) %>% 
  rename(
    Current_most_effective_method = curmethrr,
    Pregnancy_knowledge = pregknowr,
    BC_method_knowledge = totknowr,
    BC_methods_heard_about = nummthdsr,
    Age = age4grps,
    Women_protection_status = wprotect,
    Men_protection_status = mprotect,
    Protection_status = protect
  ) %>% 
  mutate(
    gender = case_when(
      gender == 1 ~ "Female",
      gender == 2 ~ "Male",
      TRUE ~ NA_character_
    ),
    race = case_when(
      race == 1 ~ "Hispanic",
      race == 2 ~ "Non-Hispanic White",
      race == 3 ~ "Non-hispanic Black",
      race == 4 ~ "Asian",
      race == 5 ~ "Other",
      TRUE ~ NA_character_
    ),
    Age = case_when(
      Age == 1 ~ "18-19",
      Age == 2 ~ "20-21",
      Age == 3 ~ "22-24",
      Age == 4 ~ "25-29",
      TRUE ~ NA_character_
    ),
    riskr = case_when(
      riskr == 1 ~ "No",
      riskr == 2 ~ "Yes",
      TRUE ~ NA_character_
    ), 
    abstonly = case_when(
      abstonly == 1 ~ "Abstinence Only",
      abstonly == 2 ~ "Abstinence Focused",
      abstonly == 3 ~ "Contraceptive Focused",
      is.na(abstonly) ~ "No sex ed/ don't know",
      TRUE ~ NA_character_
    ), 
    Protection_status = case_when(
      Protection_status == 1 ~ "Well protected",
      Protection_status == 2 ~ "Unprotected",
      Protection_status == 3 ~ "Uncertain status",
      Protection_status == 7 ~ "No Sex",
      TRUE ~ NA_character_
    ),
    Women_protection_status = case_when(
      Women_protection_status == 1 ~ "perfect pill use",
      Women_protection_status == 2 ~ "perfect condom use",
      Women_protection_status == 3 ~ "perfect depo use",
      Women_protection_status == 4 ~ "long-acting method",
      Women_protection_status == 5 ~ "skipped pills",
      Women_protection_status == 6 ~ "missed condom use",
      Women_protection_status == 7 ~ "depo late",
      Women_protection_status == 8 ~ "other method",
      Women_protection_status == 9 ~ "no method - likely at risk",
      Women_protection_status == 17 ~ "no method - unkown risk",
      Women_protection_status == 19 ~ "no sex ever/last 12 mo",
      TRUE ~ NA_character_
    ),
    Men_protection_status = case_when(
      Men_protection_status == 1 ~ "use horm/larc",
      Men_protection_status == 2 ~ "perfect condom use",
      Men_protection_status == 3 ~ "inconsistent condom use",
      Men_protection_status == 4 ~ "withdrawl",
      Men_protection_status == 5 ~ "other/unknown",
      Men_protection_status == 6 ~ "no method - risk likely",
      Men_protection_status == 7 ~ "no method - unknown risk",
      Men_protection_status == 8 ~ "no sex ever/ last 12 mo",
      TRUE ~ NA_character_
    ), 
    Current_most_effective_method = case_when(
      Current_most_effective_method == 1 ~ "long-acting alone",
      Current_most_effective_method == 2 ~ "long-acting + condom",
      Current_most_effective_method == 3 ~ "hormonal alone",
      Current_most_effective_method == 4 ~ "hormonal + condoms",
      Current_most_effective_method == 5 ~ "condoms",
      Current_most_effective_method == 6 ~ "withdrawl/NFP",
      Current_most_effective_method == 7 ~ "other/unknown",
      Current_most_effective_method == 17 ~ "no method - risk unknown",
      Current_most_effective_method == 18 ~ "no method - likely at risk",
      Current_most_effective_method == 19 ~ "no sex ever/ last 12 mo",
      TRUE ~ NA_character_
    ),
    Pregnancy_knowledge = case_when(
      Pregnancy_knowledge == 1 ~ "0-3 correct",
      Pregnancy_knowledge == 4 ~ "4 correct",
      Pregnancy_knowledge == 5 ~ "5 correct",
      Pregnancy_knowledge == 6 ~ "6 correct",
      TRUE ~ NA_character_
    ),
    BC_method_knowledge = case_when(
      BC_method_knowledge == 1 ~ "0-7 correct",
      BC_method_knowledge == 2 ~ "8-11 correct",
      BC_method_knowledge == 3 ~ "12-15 correct",
      BC_method_knowledge == 4 ~ "16-18 correct",
      BC_method_knowledge == 5 ~ "19+ correct",
      TRUE ~ NA_character_
    ),
    BC_methods_heard_about = case_when(
      BC_methods_heard_about == 1 ~ "0-6 methods",
      BC_methods_heard_about == 2 ~ "7-10 methods",
      BC_methods_heard_about== 3 ~ "11-13 methods",
      BC_methods_heard_about == 4 ~ "14-16 methods",
      TRUE ~ NA_character_
    )
  )
```   

## Visualization 1: Sexual Health Education and Risk of Unintended Pregnancy (Grouped Bar Chart)

This plot displays the percentage of participants who perceive themselves to be at risk of unintended pregnancy, grouped by the type of sexual education they received. 

```{r}
# your code here
library(ggplot2)
library(viridis)
library(dplyr)

clean_fogzone <- readRDS("data/clean_fogzone.rds")


clean_fogzone_factors <- clean_fogzone %>%
  mutate(
    abstonly = factor(
      abstonly,
      levels = c("No sex ed/ don't know", "Abstinence Only", "Abstinence Focused", "Contraceptive Focused")
    ),
    riskr = factor(riskr, levels = c("No", "Yes"))
  )

percent_data <- clean_fogzone_factors %>%
  group_by(abstonly, riskr) %>%
  summarise(n=n(), .groups="drop") %>%
  group_by(abstonly) %>%
  mutate(pct = n / sum(n) * 100)

ggplot(percent_data, aes(x = abstonly, y = pct, fill = riskr)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_text(
    aes(label = paste0(round(pct, 1), "%")),
    position = position_dodge(width = 0.9),
    vjust = -0.4,
    size = 3.5
  ) +
  scale_fill_viridis_d(
    option="D",
    name = "At Risk?",
    labels = c("No", "Yes")
  ) +
  labs(
    title = "Sexual Health Education and Risk of Unintended Pregnancy",
    x = "Type of Sexual Health Education Recived",
    y = "Percentage of Participants"
  ) +
  ylim(0,100) + 
  theme_minimal(base_size = 12) +
  theme(
    plot.margin = margin(50, 30, 30, 30),  # Adjust all margins (top, right, bottom, left)
    axis.text.x = element_text(angle = 20, hjust = 1, face = "bold", size = 10),
    axis.text.y = element_text(face = "bold"),
    plot.title = element_text(face = "bold"),
    legend.position = "top"
  )


```

## Visualization 2: Contraceptive Methods by Type of Sexual Education (Faceted Bar Chart)

This plot displays the contraceptive methods used by participants grouped by their type of sex ed.

```{r}

contraceptive_data <- clean_fogzone_factors %>%
  filter(!is.na(Current_most_effective_method)) %>%
  mutate(
    method_grouped = case_when(
      Current_most_effective_method %in% c("no method - likely at risk", "no method - risk unknown") ~ "No method",
      Current_most_effective_method %in% c("withdrawl/NFP", "other/unknown") ~ "Other/Unknown/Withdrawal",
      TRUE ~ Current_most_effective_method
    ),

    method_grouped = recode(method_grouped,
      "condoms" = "Condoms",
      "hormonal + condoms" = "Hormonal + Condoms",
      "hormonal alone" = "Hormonal",
      "long-acting + condom" = "LARC + Condoms",
      "long-acting alone" = "LARC",
      "No method" = "No Method",
      "no sex ever/ last 12 mo" = "No Sex (Last 12 Mo)",
      "Other/Unknown/Withdrawal" = "Other/Unknown/Withdrawal"
    ),
    method_grouped = factor(method_grouped),
    

    abstonly = recode(abstonly,
      "No sex ed/ don't know" = "No Sex Ed/\nDon't know",
      "Abstinence Only" = "Abstinence\nOnly",
      "Abstinence Focused" = "Abstinence\nFocused",
      "Contraceptive Focused" = "Contraceptive\nFocused"
    )
  )


contraceptive_summary <- contraceptive_data %>%
  group_by(abstonly, method_grouped) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(abstonly) %>%
  mutate(pct = n / sum(n) * 100)


ggplot(contraceptive_summary, aes(x = method_grouped, y = pct, fill = method_grouped)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_d(
    name = "Contraceptive Method",
    option = "C"
  ) +
  labs(
    title = "Contraceptive Methods by Type of Sexual Education",
    x = "Contraceptive Method",
    y = "Percentage of Participants"
  ) +
  ylim(0, 40) +
  facet_wrap(~abstonly, nrow = 1, scales = "free_x") +
  theme_minimal(base_size = 14) +
  theme(
    plot.margin = margin(t = 60, r = 30, b = 30, l = 30),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5, size = 12),
    strip.text = element_text(face = "bold", size = 8),
    legend.position = "right",
    legend.title = element_text(face = "bold", size = 8),
    legend.text = element_text(size = 8)
  )


```

## Visualization 3: Flow of Sex Ed → Risk Perception → Contraceptive Method

This diagram shows the flow of participants through the type of sex ed they receive, whether they perceive pregnancy risk, and their chosen contraceptive method. 


```{r}
library(tidyverse)
library(ggplot2)
library(ggalluvial)
library(dplyr)

sankey_data <- contraceptive_data %>%
  filter(!is.na(riskr), !is.na(abstonly), !is.na(method_grouped)) %>%
  count(abstonly, riskr, method_grouped)  

ggplot(sankey_data,
       aes(axis1 = abstonly, axis2 = riskr, axis3 = method_grouped,
           y = n)) +
  geom_alluvium(aes(fill = method_grouped), width = 0.3, alpha=0.8) +
  geom_stratum(
    width = 0.1,
    fill = "gray90",
    color = "black") +
  geom_text(
    stat = "stratum",
    aes(label = after_stat(stratum)),
    size = 3, 
    check_overlap =TRUE, 
    na.rm = TRUE,
    label_axis3 = NULL
  ) +
  scale_x_discrete(limits = c("Sex Ed Type", "Perceived Risk", "Contraceptive Method"),
                   expand = c(0.05, 0.05)) +
  scale_fill_viridis_d(option = "C", name = "Contraceptive Method") +
  labs(
    title = "Flow from Sexual Education to Risk Perception to Contraceptive Behavior",
    y = "Number of Participants"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    legend.title = element_text(size = 8, face = "bold"),
    legend.text = element_text(size = 7),
    legend.key.size = unit(0.4, "cm"),  
    legend.spacing.y = unit(0.2, "cm"), 
    legend.position = "right",
    plot.margin = margin(t = 20, r = 30, b = 20, l = 30)
  )
```

## Interpretation

Together, these three visualizations reveal how different types of sexual health education influence participants’ self-perception of pregnancy risk and their contraceptive behavior. 

The first grouped bar chart displays the proportions of self-perceived risk of unplanned pregnancy based on the type of sex education participants received. The abstinence-only group had the highest percentage of scores out of the 4 groups. Similarly, contraceptive-focused and abstinence-focused groups also had high levels of perceived risk (69% and 68.5%, respectively). In contrast, those with no sex education or who didn’t recall their education showed lower perceived risk, with only 63% identifying themselves as at risk. These results suggest that formal sex education—whether abstinence- or contraceptive-focused—may increase awareness of pregnancy risk, whereas lack of education may leave individuals with a reduced or unclear sense of personal risk.

The second faceted bar chart shows the distribution of contraceptive methods used in the past 12 months by participants, grouped by the type of sex education received. Across all education types, “No Method” and “LARC + Condoms” are among the most common categories, with contraceptive-focused and abstinence-focused groups showing slightly higher uptake of effective methods (like hormonal or dual methods). Surprisingly, even among those with contraceptive-focused education, a significant proportion still report no method use, indicating that education alone may not ensure effective behavior change. Those who received no sex education tended to cluster more around non-use or less effective methods, underscoring the importance of both access to information and behavioral support for consistent contraceptive use.

The third Sankey diagram illustrates the pathway from participants’ type of sexual health education, to their self-perceived risk of unintended pregnancy, and finally to their reported contraceptive method use. A clear pattern emerges: individuals who received contraceptive-focused education were more likely to perceive themselves at risk and to use more effective methods, such as hormonal contraception or a combination of LARC and condoms. Conversely, those with no sex education were more likely to perceive no risk and use no method. The diagram highlights how sex education influences both awareness of pregnancy risk and subsequent contraceptive behavior, reinforcing the idea that comprehensive, contraceptive-based education better equips individuals to recognize risk and make informed reproductive health decisions.

## Limitations

This data is subject to several limitations. First, all observations rely on self-reported information about sexual health knowledge and behaviors, which introduces the potential for recall bias and social desirability bias. Participants may have overestimated their safe sex practices or under-reported riskier behaviors. Second, while the dataset includes many helpful variables, it lacks information about where participants received their sexual health education. Knowing the state or region could enable geographic visualizations or an interactive Shiny app to explore the effects of state-specific curricula—an insight that could support policy arguments for more comprehensive sex education in public schools. Finally, the data was collected in 2009 and may not reflect more recent policy changes, cultural shifts, or behavioral trends among young people.


## Conclusion

Conclusively, these results suggest that the type of sex ed that participants receive can significantly influence both their self-perceived risk of pregnancy and their contraceptive choices. Comprehensive or contraceptive-focused education appears to foster greater risk awareness and promote the use of more effective contraceptive methods, while lack of education may contribute to lower perceived risk and less protective behavior. As a next step, these insights can inform the development of sex education curricula that not only provide accurate and scientifically up-to-date information, but also actively engage students in understanding personal risk and decision-making. Future research should explore how these patterns differ by age, gender, or cultural context, and evaluate the long-term impact of education type on actual pregnancy outcomes.

## References

1. Reference 1: Raw Data file 
Frost, Jennifer J. 2009 National Survey of Reproductive and Contraceptive Knowledge: Guttmacher 2009 Survey of contraceptive knowledge  Fog Zone .dta. Ann Arbor, MI: Inter-university Consortium for Political and Social Research [distributor], 2022-03-31. https://doi.org/10.3886/E164142V1-146384
https://www.openicpsr.org/openicpsr/project/164142/version/V1/view?path=/openicpsr/164142/fcr:versions/V1/Guttmacher-2009-Survey-of-contraceptive-knowledge--Fog-Zone-.dta&type=file

2. Reference 2: Codebook for raw data 
Frost, Jennifer J. 2009 National Survey of Reproductive and Contraceptive Knowledge: Guttmacher 2009 Survey of contraceptive knowledge  Fog Zone  user guide and codebook.pdf. Ann Arbor, MI: Inter-university Consortium for Political and Social Research [distributor], 2022-03-31. https://doi.org/10.3886/E164142V1-131024
https://www.openicpsr.org/openicpsr/project/164142/version/V1/view?path=/openicpsr/164142/fcr:versions/V1/Guttmacher-2009-Survey-of-contraceptive-knowledge--Fog-Zone--user-guide-and-codebook.pdf&type=file#

3. Reference 3: ICPSR project link
Frost, Jennifer J. 2009 National Survey of Reproductive and Contraceptive Knowledge. Ann Arbor, MI: Inter-university Consortium for Political and Social Research [distributor], 2022-03-31. https://doi.org/10.3886/E164142V1
https://www.openicpsr.org/openicpsr/project/164142/version/V1/view


